---
title: "survey_data_analysis"
author: "Keith Karani"
format: html
editor: visual
---

## Financial well-being survey data analysis

This analysis focuses on data from the National Financial Well-Being Survey to learn more about how a wide range of factors relate to consumers’ financial well-being in 2017 by the Consumer Financial Protection Bureau (cfpb).

```{r}
#| echo: fenced
# load packages to use
library(dplyr)
library(tidyr)
library(ggplot2)

```

#### Import data via url

```{r}
# import the data

fwb_data <- read.csv("https://www.consumerfinance.gov/documents/5614/NFWBS_PUF_2016_data.csv")

View(fwb_data)

```

Our analysis will focus on responses on individuals in the lower income households bracket i.e. households that earn less than \$50,000. To achieve this, we are guided by the data dictionary available in the public use file cookbook [Public Use File Codebook](https://www.consumerfinance.gov/documents/5586/cfpb_nfwbs-puf-codebook.pdf "National Financial WellBeing Survey Public Use File Codebook")

```{r}
# subset respondents who earn less than $50,000 income

low_inc <- fwb_data |> 
  filter(PPINCIMP <= 4)


# let us validate what our subset achived by observing the household income variable
table(low_inc$PPINCIMP)



# To narrow down our analysis we will focus on three variable respondents gender, respondents household size and the respondents income

low_income_ndwn <- low_inc |> 
  select(PPGENDER, PPHHSIZE, PPINCIMP, FWBscore, finalwt)

```

#### creating a binary variable

we will achieve this by focusing on the *Household Size* variable. First, we use the data dictionary available in the public use file cookbook [Public Use File Codebook](https://www.consumerfinance.gov/documents/5586/cfpb_nfwbs-puf-codebook.pdf "National Financial WellBeing Survey Public Use File Codebook") to look into the interested variable name in this case *Household Size*.

```{r}

low_income_ndwn$smaller_house <- ifelse(low_income_ndwn$PPHHSIZE <= 2, 1, 0)

View(low_income_ndwn)


```

#### Renaming the levels of a categorical variable

Our focus will be on the PPGENDER variable which contains observations of gender of respondents in the survey. After referencing the data dictionary I observed that value 1 represents male while value 2 represents female gender. With this in mind we can use the recode() function from dplyr to achieve our objective

```{r}
# recode 
low_income_ndwn$PPGENDER <- recode(low_income_ndwn$PPGENDER, "1" = "Male", "2" = "Female")

#print(low_income_ndwn)

table(low_income_ndwn$PPGENDER)
```

#### How can you create a new categorical variable

In our survey data let us create a variable based on the gender variable and smaller house variable

```{r}
low_income_ndwn$gen_small_house <- if_else(low_income_ndwn$PPGENDER == "Female" & low_income_ndwn$smaller_house == 1, "Female living in a small house",
                                         if_else(low_income_ndwn$PPGENDER == "Female" & low_income_ndwn$smaller_house == 0, "Female living in a large house",
                                                 if_else(low_income_ndwn$PPGENDER == "Male" & low_income_ndwn$smaller_house == 1, "Male living in a small house",
                                                         "Male living in a large house")))

```

Let us create a statistical summary table to estimate their financial wellbeing score abbreviated by FWBscore in the reference document whereby:

A higher score denotes more satisfaction with the finances and a lower score denotes the vice versa.

```{r}
# these values are refernced from the documentation
# recode the levels of the income variable
low_income_ndwn$PPINCIMP <- recode(low_income_ndwn$PPINCIMP,
                                   "1" = "less than $20,000",
                                   "2" = "$20,000 to $29,000",
                                   "3" = "$30,000 to $39,000",
                                   "4" = "$40,000 to $49,000"
                                   )
```

```{r}
summ_stat_tbl <- low_income_ndwn |> 
  group_by(gen_small_house, PPINCIMP) |> 
  summarise(
    average_FWBscore = round(mean(FWBscore), digit = 2),
    average_FWBscore_weighted = round(weighted.mean(FWBscore, finalwt), digits = 2),
    median_FWBscore = round(median(FWBscore), digits = 2),
    SD_FWBscore = round(sd(FWBscore), digits = 2)
  )

summ_stat_tbl


```

#### visualize the results from our summary table

```{r}

ggplot(summ_stat_tbl, aes(x = gen_small_house, y = average_FWBscore_weighted)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    x = " ",
    y = "average_FWBscore"
  ) +
  theme_minimal()


```

We can improve our visualization further buy having a grouped bar plot by working on the annual financial income

```{r}
ggplot(summ_stat_tbl, aes(x = gen_small_house, y = average_FWBscore_weighted, fill = PPINCIMP)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(
    title = "Avg financial well being score ",
    x = " ",
    y = "average financial wellbeing score", fill = "Household income"
  ) +
  theme_light()

```
